Likelihoods for a general class of ARGs under the SMC

### 2025-02-21 paper.tex
Pass through disco

Fig~\ref{fig:lik-surface} also shows results for ARGs inferred from the 
simulated data using tsinfer and tsdate~\citep{wohns_unified_2022},
using default parameters for tsinfer, and the true values of mutation
rate and $N_e$ for tsdate.
In the first, we use the default output of tsinfer 
(2,148 nodes; 5,589 edges; 1,307 trees; 3,039 mutations).
This ARG has an average of 192 unary nodes, 120 binary nodes, and 32 nodes
with three or more children per tree. Tsinfer does not directly estimate
recombination events; the breakpoints between trees are the consequence of
switches between parental haplotypes in the 
\cite{li2003modeling} copying process. Information about 
recombination events in the returned ARG is therefore quite imprecise,
and likely to contain incorrect and contradictory details if one were to 
attempt to reconstruct the exact subtree-prune-and-regraft 
moves~\citep[e.g.][]{rasmussen2023espalier}. Nonetheless, in
this toy example at least, our likelihood function performs well
on the output of tsinfer, with the minimum close to the true
parameter value. We also show the likelihood curve obtained when we 
fully simplify the output of tsinfer 
(1,865 nodes; 7,511 edges; 1,252 trees; 3,039 mutations)
before dating. We can see that, as with the case of the true data,
removing all the unary nodes results in a significant bias
away from the true value of the recombination rate parameter.

@@

\section*{Discussion}

The key contribution of this paper is to rederive a classical
ARG likelihood function in a way that supports the sort of partial information
provided by some modern ARG inference tools,
using a novel backwards-time derivation of the SMC.
The underlying process that we reason about probabilistically is still
composed of events affecting lineages, but, because we
do not necessarily observe or estimate these events,
we integrate over the possible timings implied by the information
in the input ARG.
Our method can compute a likelihood under the SMC for a very general class of 
ARG~\citep{wong_general_2023}, 
a significant increase in flexibility over the 
strict input requirements for the classical Kuhner-Yamato-Felsenstein 
formulation. 
We have shown via some illustrative examples that the formulation is 
remarkably robust to incomplete information, tolerating the loss 
of a significant fraction of the internal nodes in the ARG to minimal effect.
% I think we have to grasp this nettle or we're inviting the reviewers to 
% give us a bunch of "what if you tried this" bits of extra homework.
While there are many different potential applications to concrete 
inferential problems, the details are nontrivial and best consigned to future work.

A major advantage of our approach is that it does not require
us to explicitly identify common ancestor and recombination events 
in the input. Indeed, 
we can compute a likelihood for any ARG in \texttt{tskit} format,
potentially bridging the gap between population genetic theory and the 
varied structures inferred by current methods~\citep{wong_general_2023}.

there are some caveats. By treating each edge independently,
our algorithm will still produce a value for ARGs with
the various degrees of completeness to which inheritance
is documented by the different ARG inference methods.
% : e.g., polytomies, absence
% of locally unary coalescence nodes, unpreserved identify of internal nodes
% across local trees, etc.
% However, many of these imprecisions violate the implicit
% assumptions of the likelihood computation.
For example, we assume the 
presence of locally unary nodes to correctly identify the
full extent of all lineages going back through time 
and therefore recombination events, but this information
may be incomplete (e.g., the full simplifed ARG in Fig.~\ref{fig:lik-surface})
or even incorrect.
% Without these unary nodes, our algorithm will detect a recombination
% event along both coalescing lineages in case of a (sub)tree height changes along
% the ARG as both edges would switch parent nodes. 
Locally unary nodes are not a common aspect of current ARG 
research~\citep{wong_general_2023}, but 
are estimated by tsinfer, and can be imputed to high
degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
Determining the effects of various forms of \emph{incorrect} topology on
these likelihoods requires further research.

[Something about polytomies]

Our results, while preliminary, suggest that ``full ARGs'', in which 
each recombination event is precisely pinpointed in time, 
contain a significant degree of redundancy in terms of capturing 
information about the generative process. 
It is possible, therefore, that explicitly estimating recombination 
events is not necessary to sample ARGs under the SMC (as ARGweaver
and SINGER do), potentially
leading to further improvements in the scalability of such methods.
Another exciting prospect for ARG inference is the development of 
``hybrid'' inference methods, where hueristic methods such as 
tsinfer or ARG-Needle are used to infer
the details of the recent past in huge datasets such as 
UK Biobank, and passing over to a rigorous model-based
approach for the ancient past.
This is similar in spirit to the widely-used approach 
of simulating complex dynamics of the recent past using 
detailed but slow models and using faster coalescent simulations
in the more distant past, when the assumptions required 
are more justified~\citep{bhaskar_distortion_2014,
kelleher_efficient_2018,haller_tree_2018,nelson_accounting_2020}.
% Maybe we shouldn't bother saying this, but I can't resist the 
% opportunity to say "hey, you should be using tskit!"
% It feels like some PIs aren't getting this message, so need to have
% it spelled out.
Just as forward and backward simulations are useful in different contexts
and can be combined to great effect,
large-scale hueristic and MCMC sampling inference methods have their 
applications, and there may be synergy in their combination. 
Practical combination of methods will require cooperation at the 
data structure level, and the interoperability provided by the 
\texttt{tskit} library [something good].

>>>>

Fig~\ref{fig:lik-surface} also shows results for ARGs inferred from the 
simulated data using tsinfer and tsdate~\citep{wohns_unified_2022},
using default parameters for tsinfer, and the true values of mutation
rate and $N_e$ for tsdate.
In the first, we use the default output of tsinfer 
(2,148 nodes; 5,589 edges; 1,307 trees; 3,039 mutations).
This ARG has an average of 192 unary nodes, 120 binary nodes, and 32 nodes
with three or more children per tree. Tsinfer does not directly estimate
recombination events; the breakpoints between trees are the consequence of
switches between parental haplotypes in the 
\cite{li2003modeling} copying process. Information about 
recombination events in the returned ARG is therefore quite imprecise,
and likely to contain incorrect and contradictory details if one were to 
attempt to reconstruct the exact subtree-prune-and-regraft 
moves~\citep[e.g.][]{rasmussen2023espalier}. Nonetheless, in
this toy example at least, our likelihood function performs well
on the output of tsinfer, with the minimum close to the true
parameter value. We also show the likelihood curve obtained when we 
fully simplify the output of tsinfer 
(1,865 nodes; 7,511 edges; 1,252 trees; 3,039 mutations)
before dating. We can see that, as with the true data,
removing all the unary nodes results in a significant bias
away from the true value of the recombination rate parameter.

@@

\section*{Discussion}
The key contribution of this paper is to rederive a classical
ARG likelihood function in a way that supports the sort of partial 
and imprecise information
provided by some modern ARG inference tools,
using a novel backwards-time derivation of the SMC.
The underlying process that we reason about probabilistically is still
composed of events affecting lineages, but, because we
do not necessarily observe or estimate these events,
we integrate over the possible timings implied by the information
in the input ARG.
Our method can compute a likelihood under the SMC for a very general class of 
ARG~\citep{wong_general_2023}, 
a significant increase in flexibility over the 
strict input requirements for the classical Kuhner-Yamato-Felsenstein 
formulation. 
We have shown via some illustrative examples that the formulation is 
robust to incomplete and imprecise information, tolerating the loss 
of a significant fraction of the internal nodes in the ARG to minimal effect.
We have also shown that the likelihood function performs well 
(at least in our toy example) on ARGs inferred by tsinfer+tsdate,
successfully capturing information about latent recombination events
from imprecise and noisy data.
Locally unary nodes play a key role, and although not
currently a common focus of ARG research~\citep{wong_general_2023}, 
they are estimated by tsinfer and can be imputed to high a
degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
% I think we have to grasp this nettle or we're inviting the reviewers to 
% give us a bunch of "what if you tried this" bits of extra homework.
Fully characterising the properties of the likelihood on the output 
of tsinfer and other methods is beyond the scope of this work.
Similarly, while there are many different potential applications to concrete 
inferential problems, the details are nontrivial and must be
consigned to future work.

% Structural uncertainty

% A major advantage of our approach is that it does not require
% us to explicitly identify common ancestor and recombination events 
% in the input. Indeed, 
% we can compute a likelihood for any ARG in \texttt{tskit} format,
% potentially bridging the gap between population genetic theory and the 
% varied structures inferred by current methods~\citep{wong_general_2023}.

% there are some caveats. By treating each edge independently,
% our algorithm will still produce a value for ARGs with
% the various degrees of completeness to which inheritance
% is documented by the different ARG inference methods.
% % : e.g., polytomies, absence
% % of locally unary coalescence nodes, unpreserved identify of internal nodes
% % across local trees, etc.
% % However, many of these imprecisions violate the implicit
% % assumptions of the likelihood computation.
% For example, we assume the 
% presence of locally unary nodes to correctly identify the
% full extent of all lineages going back through time 
% and therefore recombination events, but this information
% may be incomplete (e.g., the full simplifed ARG in Fig.~\ref{fig:lik-surface})
% or even incorrect.
% % Without these unary nodes, our algorithm will detect a recombination
% % event along both coalescing lineages in case of a (sub)tree height changes along
% % the ARG as both edges would switch parent nodes. 
% Locally unary nodes are not a common aspect of current ARG 
% research~\citep{wong_general_2023}, but 
% are estimated by tsinfer, and can be imputed to high
% degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.
% Determining the effects of various forms of \emph{incorrect} topology on
% these likelihoods requires further research.

% [Something about polytomies]

Our results, while preliminary, suggest that ``full ARGs'', in which 
each recombination and common ancestor event is precisely estimated,
contain a significant degree of redundancy in terms of capturing 
information about the generative process. 
It is possible, therefore, that explicitly estimating 
\emph{all} such events is not necessary to sample ARGs under the 
SMC (as ARGweaver and SINGER do). 
A combined approach, in which the uncertainty around the
ordering of events that are 
weakly informed by the data is encoded structurally in the 
graph and other forms of uncertainty captured by sampling from 
the posterior, may yield further improvements in 
scalability for statistically rigorous inference.
% Signicant work would be needed, however, to make 
% downstream analysis robust to such stuctural uncertainty;
% polytomies, for example, are often not
Another exciting prospect for ARG inference is the development of 
``hybrid'' inference methods, where hueristic methods such as 
tsinfer or ARG-Needle are used to infer
the details of the recent past in huge datasets such as 
UK Biobank, passing over to a rigorous model-based
approach for the ancient past.
This is similar in spirit to the widely-used approach 
of simulating complex dynamics of the recent past using 
detailed but slow models and using faster coalescent simulations
in the distant past when the assumptions required 
are more justified~\citep{bhaskar_distortion_2014,
kelleher_efficient_2018,haller_tree_2018,nelson_accounting_2020}.
% Maybe we shouldn't bother saying this, but I can't resist the 
% opportunity to say "hey, you should be using tskit!"
% It feels like some PIs aren't getting this message, so need to have
% it spelled out.
Just as forward and backward simulations are useful in different contexts
and can be combined to great effect,
large-scale hueristic and MCMC sampling inference methods have their 
applications, and there may be synergy in their combination. 
% This is pretty weak, tbh.
The practical details of ``handing over'' between methods
will require cooperation and precise communication;
the tskit library, with its well-defined data model and flexible
metadata support, is the ideal basis for this.
